---
title: "Lab 7: Sampling Distributions of Sample Proportions "
author: "Name"
date: "08/06/20"
output: oilabs::lab_report
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
ames <- read_csv("http://bryandmartin.github.io/STAT311/oiLabs/Week7/ames.csv")
```

* * *

### Exercise 1: 

```{r}


```

### Exercise 2:

```{r}

```

**written answer**.

### Exercise 3: 

```{r}

```

**written answer**


### Exercise 4: 

```{r}

```

**written answer: what happens when you run code over and over. **

### Exercise 5:


### Exercise 6:

```{r}
samp1 <- sample(air, size=20, replace=TRUE)
samp2 <- sample(air, size=50, replace=TRUE)
samp3 <- sample(air, size=200, replace=TRUE)
### Print p1
### Print p2
### Print p3

```

**written answer**

### Exercise 7: 

```{r}
set.seed(111)
ps_20 <- replicate(100000, mean(sample(air, size=20, replace=TRUE)))
```

**describe the plot**

### Exercise 8

```{r}

```

**written answer**


### Exercise 9

```{r}

```

**written answer**

### Exercise 10

```{r}
set.seed(111)
phats_20 <- replicate(100000, mean(sample(air, size=20, replace=TRUE)))
### Fill in for size 50
### Fill in for size 200
```

### Exercise 11
```{r}
## Two histograms
```

**Written explanation**

### Exercise 12

**Written answer**

### Exercise 13

**Written answer**

### Exercise 14

```{r}
ggplot(data = NULL, aes(x = ps_200)) +
        geom_blank() +
        geom_histogram(bins=30,aes(y = ..density..)) +
        stat_function(fun = dnorm, args = c(mean = pi, sd = sqrt((pi)*(1-pi)/200)), col = "tomato")
```

**Written answer**

### Exercise 15

```{r}
## Include lower and upper bound code
## And calculations
```

**Report final answer**. 


