---
title: "Lab 4: Exploring Categorical Data"
author: "Name"
date: "07/07/2020"
output: oilabs::lab_report
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
cdc <- read.table("http://www.stat.uchicago.edu/~yibi/s220/labs/data/cdc.dat", header=TRUE)
cdc <- cdc %>% mutate(genhlth = ordered(genhlth, levels=c("poor", "fair", "good", "very good", "excellent"))) 
```

* * *

### Exercise 1: 

```{r}
# Include the relative frequency table here. 
```

(write you answer here- don't forget to submit on Canvas as well).

### Exercise 2:

```{r}
# Create a barplot
```


### Exercise 3:

```{r}
# Include your code to add the BMI variable
# Include your code to find the mean BMI
```

### Exercise 4:

```{r}
# Make the boxplot
```

(describe the relationship)

### Exercise 5

```{r}
# Create table here to compute your answer
```

(write your answer here)

### Exercise 6

```{r}
# include the joint frequency table
```

(write your answer here. )

### Exercise 7

```{r}
# Include your mosaic plot
```
(write your answer)

### Exercise 8

```{r}
# Include the relevant table
```
(write your sentence here)

### Exercise 9

(write your answer here- additional code not required)

### Exercise 10

```{r}
# Make the relevant plot
```
(write your answer)


### Exercise 11

```{r}
# Make the new plot
```
 (comment)


