---
title: "Lab 3: Exploring Numerical Data"
author: "Name"
date: "Date of lab session"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
## Include the command to load the data here!!!!
```

* * *

### Exercise 1: 

(Write your answer for exercise 1 here. This exercise doesn't have any code associated with it) 

### Exercise 2: 

```{r}
# Insert code for exercise 2 here 
# (make the sfo_feb_flights dataset and count the number of observations)
```

(write your final answer here)

### Exercise 3: 


```{r}
# Insert code for exercise 3 here 
# (make the histogram) 
```

Write the answer to exercise 3 here (describe the distribution)

### Exercise 4: 

```{r}
# Insert code for exercise 4 here 
# (calculate mean and median)
```

Write the answer to exercise 4 here (compare mean and median, interpret difference). 


### Exercise 5:

```{r}
# Insert code for exercise 5 here 
# (group by, summarize by)
```

Write the answer to exercise 5. Don't forget to also submit the answer on Canvas.


### Exercise 5:


```{r}
# Insert code for exercise 6 here 
# (group by, summarize by)
```

Write the answer to exercise 6. 

### Exercise 7:

Write the answer to exercise 7.

```{r}
# Insert code for exercise 7 here 
# (group by, summarize by)
```


### Exercise 8:

Write the answer to exercise 8 (describe the relationship)

```{r}
# Insert code for exercise 8 here 
# (make the scatterplot)
```


### Exercise 9

Write your explanation for exercise 9 here. 

```{r}
# Back up the answer to exercise 9 with relevant code
```





